Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-01-10 ParisEM.tex
New minor corrections - All sections

The exact computation of these expectations is usually not possible for general partially observed diffusions. In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights. Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly. Many discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or a Gaussian based approximation using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}. Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein. However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see \cite{delmoral:jacod:protter:2001}. To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.  In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.

The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
Approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$. Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass. Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e. using the observations as they are received) smoothed expectations of additive functionals. Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.

In this paper, instead of using fixed-lag techniques which introduce a systematic additional bias in the approximation of conditional expectations, we extend the use of  PaRIS algorithm to partially observed SDE. The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles. The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by bounded unbiased estimates. We also propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates, and to extend the range of diffusion processes for which SMC methods can be applied. To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}. These new GPE estimators use a an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.

\section{Random weight PaRIS algorithm}
In this paper, $(X_t)_{t\ge 0}$ is a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion. It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function. 
A popular choice of $A$  is $A = \log \pi/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$. See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi$. SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.

@@

The distribution of $X_0$ has a density with respect to $\lambda$ given by $\chi$.
For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$. 

@@

An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.  

@@

The PaRIS algorithm combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. 

@@

Assume also that the model is such that there exists $\Lambda_k$ satisfying  $\mathrm{sup}_x\eqsp \phi_{k}(x)\le \Lambda_{k}$. These assumptions might seem somehow restrictive but are necessary to define the transition density estimator in the linear case. The alternative given by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016} when the instrumental drift is constant does not assume any upper bound for $\phi_{k}$ but requires to sample skeletons of a Brownian bridge given its minimum/maximum between $t_k$ and $t_{k+1}$. To the best of our knowledge, there is no closed from formula to sample skeletons of an Ornstein-Uhlenbeck bridges given its extremal values on an interval. Then,

@@

An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]$. 

@@

&\hspace{-3cm}= \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}} \left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]\eqsp,\\

@@

where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$. The resulting Ozaki In PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.

@@

\subsection{Constant approximation of the drift}
\label{sec:constantdrift}
In the case where the function $\phi_k$ cannot be upper bounded when $\widetilde{\alpha}_k$ is set as a linear approximation of $\alpha$, \cite{giesecke:schwenkler:2016} suggests to set $\widetilde{\alpha}_k$ as a constant value $\widetilde{\alpha}_k= \rho_k$. The GPE of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} is a special case with $\rho_k = 0$. In this case, $H_k$ may be chosen as $H_k: x\mapsto A(x) - \rho_k x$. Then, following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,olsson:strojby:2011}, it is enough to sample a pair $(L_k,U_k)$ of random variables and a Brownian bridge $\mw\sim \mathbb{W}_k^{X_{k+1}}$ (or a Brownian bridge with a constant drift $\rho_k$) such that, for all $t_k\le t\le t_{k+1}$,
\[
L_k\le \phi_k(\mw_s)\le U_k\eqsp.
\]
Sampling the bridge conditionally on its minimum and maximum is possible using Bessel bridges as noted by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}. Then, the second term of \eqref{eq:unbiased:q} is replaced by:
\[
\mathrm{exp}\left\{-U_{k}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(U_{k}-\phi_{k}(\mw_{U_j})\right)\eqsp.
\]
Here again, the choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$, see Section~\ref{sec:exp}.

\section{Convergence results}
Consider the following assumptions.
\begin{hypA}
\label{assum:boundmodel}
\begin{enumerate}[(i)]
\item For all $k \geq 0$ and  all $x\in \mathbb{R}^d$, $g_{k}(x) >0$.
\item $\underset{k\geq 0}{\sup}|g_{k}|_{\infty} < \infty$.
\end{enumerate}
\end{hypA}

@@

@@

\section{Proofs}
\label{sec:append:proofs}
\begin{proof}[Proof of Lemma~\ref{lem:AR:unbiased}]
%Let $\tau$ be the first time  draws are accepted in the accept-reject mechanism. For all $\ell\ge 1$, write
%\[
%\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})/\sigma_+\right\}\eqsp.
%\]
%
%Let $h$ be a function defined on $\{1,\ldots,N\}$,
%\begin{align*}
%\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(I_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
%& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
%& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
%&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_m},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
%& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
%& = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
%& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
%&= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
%\end{align*}
%which concludes the proof.
The goal is to simulate realisations of  a discrete random variable $X$ having the following target distribution:
\mP(X=\ell)= p_X(\ell)=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{\ell=1}^N\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})},~~\ell\in \{1,\dots,N\}
for some $\xi_{k+1}$ that has no importance in this result. We assume $\omega_k$ are normalized weights, i.e. $\sum_\ell \omega_k^l=1$. The candidate will therefore be sampled from $Y$, a r.v. with the following distribution:
\mP(Y=\ell)=p_Y(\ell)= \omega_k^\ell,~~\ell\in \{1,\dots,N\}
Unfortunately, the $q_k$ function is unknown. Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f. $p_\zeta$.  We have therefore:
$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)\right]=q(\xi_k^\ell,\xi_{k+1})$$
Moreover, we impose the following property over this estimator, we want that:
$$\exists \sigma_+\in \mathbb{R}\text{ such that }\forall x,y,~~\hat{q}_k(x,y,\zeta)\leq \sigma_+$$
To simulate a random variable $X\sim p_X$, we propose the following algorithm:
\begin{enumerate}
\item Sample independently $\ell\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
\item \textbf{If} 
$$u \leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},z)}{\sigma_+},$$
set $x=\ell$. \textbf{Else}, return to 1).

>>>>

The exact computation of these expectations is usually not possible in the case of partially observed diffusions. In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights. Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly. Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}. Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein. However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}. To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.  In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.

The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$. Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass. Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e. using the observations as they are received) smoothed expectations of additive functionals. Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.

In this paper, instead of using fixed-lag techniques which introduce a systematic additional bias in the approximation of conditional expectations, we extend the use of  the PaRIS algorithm to partially observed SDE. The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles. The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by bounded unbiased estimates. We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied. To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}. These new GPE estimators use an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.

The paper is organized as follows. Section~\ref{sec:rwparis} describes the random weight PaRIS algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states. Then, Section~\ref{sec:PaRIS:SDE} details the application of this algorithm when the transition density may be approximated using a GPE based on the Ozaki discretization or a GPE based on a constant approximation of the drift function. The latter is a straightforward generalization of \cite{fearnhead:papaspiliopoulos:roberts:2008}. In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. All proofs are postponed to Appendix~\ref{sec:append:proofs}.

\section{Random weight PaRIS algorithm}
\label{sec:rwparis}
$(X_t)_{t\ge 0}$ is defined as a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion. It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function. 
A popular choice of $A$  is $A = \log \pi/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$. See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi$. SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.

@@

The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$. 

@@

%An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.  

@@

It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. 

@@

Assume also that the model is such that there exists $\Lambda_k$ satisfying  $\mathrm{sup}_x\eqsp \phi_{k}(x)\le \Lambda_{k}$. These assumptions might seem somehow restrictive but are necessary to define the transition density estimator in the linear case. The alternative given by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016} when the instrumental drift is constant does not assume any upper bound for $\phi_{k}$ but requires to sample skeletons of a Brownian bridge given its minimum/maximum between $t_k$ and $t_{k+1}$. To the best of our knowledge, there is no closed from formula to sample skeletons of an Ornstein-Uhlenbeck bridge given its extremal values on an interval. Then,

@@

An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating the expectation in the previous equation. 

@@

@@

where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$. The resulting Ozaki in PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.

@@

\subsection{Constant approximation of the drift}
\label{sec:constantdrift}
In the case where the function $\phi_k$ cannot be upper bounded when $\widetilde{\alpha}_k$ is set as a linear approximation of $\alpha$, \cite{giesecke:schwenkler:2016} suggests to set $\widetilde{\alpha}_k$ as a constant value $\widetilde{\alpha}_k= \rho_k$. The GPE of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} is a special case with $\rho_k = 0$. In this case, $H_k$ may be chosen as $H_k: x\mapsto A(x) - \rho_k x$. Then, following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,olsson:strojby:2011}, it is enough to sample a pair $(L_k,U_k)$ of random variables and a Brownian bridge $\mw\sim \mathbb{W}_k^{X_{k+1}}$ (or a Brownian bridge with a constant drift $\rho_k$) such that, for all $t_k\le t\le t_{k+1}$,
\[
L_k\le \phi_k(\mw_s)\le U_k\eqsp.
\]
Sampling the bridge conditionally on its minimum and maximum is possible using Bessel bridges as noted by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}. Then, the second term of \eqref{eq:unbiased:q} is replaced by:
\[
\mathrm{exp}\left\{-U_{k}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(U_{k}-\phi_{k}(\mw_{U_j})\right)\eqsp.
\]
Here again, the choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$.

\section{Convergence results}
\label{sec:convergence}
Consider the following assumptions.
\begin{hypA}
\label{assum:boundmodel}
\begin{enumerate}[(i)]
\item For all $k \geq 0$ and  all $x\in \mathbb{R}^d$, $g_{k}(x) >0$.
\item $\underset{k\geq 0}{\sup}|g_{k}|_{\infty} < \infty$.
\end{enumerate}
\end{hypA}

@@

\begin{proposition}
\label{prop:filter}
Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold. Then, for all $0\le k\le n$,
there exist $b_k, c_k>0$ such that for all $N\ge 1$, all $\varepsilon > 0$, and all measurable function $h$ on $\mathsf{X}$,
\begin{equation*}
\mathbb{P}\left( \left|\phi_k^N[h] - \phi_k\left[h\right]\right| \geq \varepsilon \right) \leq b_k \mathrm{exp}\left(-c_k N \epsilon^2\right)\eqsp.
\end{equation*}
\end{proposition}

\begin{proof}
\textcolor{red}{On l'ecrit peut-etre... }
The proof is a straightforward extension of \cite{douc:garivier:moulines:olsson:2011}. 
\end{proof}

@@

\section{Proofs}
\label{sec:append:proofs}
\begin{proof}[Proof of Lemma~\ref{lem:AR:unbiased}]
Let $\tau$ be the first time  draws are accepted in the accept-reject mechanism. For all $\ell\ge 1$, write
\[
\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})/\sigma_+\right\}\eqsp.
\]

Let $h$ be a function defined on $\{1,\ldots,N\}$,
\begin{align*}
\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(I_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_m},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
&= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
\end{align*}
which concludes the proof.

%The goal is to simulate realisations of  a discrete random variable $X$ having the following target distribution:
%\begin{equation}
%\mP(X=\ell)= p_X(\ell)=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{\ell=1}^N\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})},~~\ell\in \{1,\dots,N\}
%\end{equation}
%for some $\xi_{k+1}$ that has no importance in this result. We assume $\omega_k$ are normalized weights, i.e. $\sum_\ell \omega_k^l=1$. The candidate will therefore be sampled from $Y$, a r.v. with the following distribution:
%\begin{equation}
%\mP(Y=\ell)=p_Y(\ell)= \omega_k^\ell,~~\ell\in \{1,\dots,N\}
%\end{equation}
%Unfortunately, the $q_k$ function is unknown. Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f. $p_\zeta$.  We have therefore:
%$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)\right]=q(\xi_k^\ell,\xi_{k+1})$$
%Moreover, we impose the following property over this estimator, we want that:
%$$\exists \sigma_+\in \mathbb{R}\text{ such that }\forall x,y,~~\hat{q}_k(x,y,\zeta)\leq \sigma_+$$
%To simulate a random variable $X\sim p_X$, we propose the following algorithm:
%\begin{enumerate}
%\item Sample independently $\ell\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
%\item \textbf{If} 
%$$u \leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},z)}{\sigma_+},$$
%set $x=\ell$. \textbf{Else}, return to 1).
